---
title: "Count Active Events by Hour Interval"
author: "Donnie Minnick"
date: "`r Sys.Date()`"
output:
  md_document:
    variant: markdown_github
---

## Problem

You need to count active events in a manner that require a solution for counting by hour interval, accounting for events that overlap intervals.

## Solution

This solution is implemented as a  function that count active events for each hour, considering the start and end hours of those events. The function accounts for events that overlap multiple hour intervals to avoid counting them more than once.

This solution assumes events start and end within a single 24-hour day.

#### Load Libraries

This solution uses the `dplyr` and `knitr` libraries.  The code chunk below checks for installation and, if needed, will install the libraries and load them.  The `ggplot` library is used to generate the plot at the end of this example.

```{r load_libraries, message = FALSE}
packages <- c("dplyr", "ggplot2", "knitr")

installed_packages <- packages %in% rownames(installed.packages())

if(any(installed_packages == FALSE)) {
  
  install.packages(packages[!installed_packages])
  
}

invisible(lapply(packages, library, character.only = TRUE))
```

#### Create Example

To demonstrate how to count active events by hour interval, I'll create an example data frame.  

This data frame has four variables, an event id and integer values for start hour, end hour and duration, which is calculated by subtracting the start hour from the end hour.

```{r create_example, messages = FALSE}
events <- data.frame(event_id = c("A", "B", "C", "D", "E", "F", "G", "H", "I"),
                     start_hour = c(0, 1, 2, 3, 4, 5, 6, 7, 13),
                     end_hour = c(1, 3, 5, 7, 9, 11, 13, 15, 17))

events <- events %>%
    mutate(duration = end_hour - start_hour)
```

With the exception of the first event, all the events in this data frame will overlap at least one interval, and there will be hours in the day in which there are no active events.  The data frame looks like this.

```{r show_example, messages = FALSE}
kable(events,
      col.names = c("Event", "Start Hour", "End Hour", "Duration"),
      caption = "Events Data Frame",
      format.args = list(big.mark = ","),
      align = c("l", "r", "r", "r"))
```

#### Count Function

Explain...

```{r count_function, messages = FALSE}
count_active_events <- function(current_hour, start_hour, end_hour) {
  
  sum(start_hour <= current_hour & end_hour >= current_hour)

}
```

#### Generate Hourly Counts

Use function to create counts table...

```{r generate_event_counts, message = FALSE}
hours <- 0:23

hourly_counts <- data.frame(hour_of_day = hours)

hourly_counts <- hourly_counts %>%
  rowwise() %>%
  mutate(active_event_count = count_active_events(hour_of_day, events$start_hour, events$end_hour))
```

Resulting table...

```{r show_event_counts, message = FALSE}
kable(hourly_counts,
      col.names = c("Hour", "Active Event Count"),
      caption = "Active Events by Hour Interval",
      format.args = list(big.mark = ","),
      align = c("l", "r"))
```

## Discussion

This solution is useful when you want to visualize active events over the course of a day.

```{r plot_active_events, messages = FALSE}
ggplot(hourly_counts, aes(x = hour_of_day, y = active_event_count)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(title = "Active Events by Hour Interval",
       subtitle = "Activity peaks during the hours of 7:00-9:00 am.",
       x = "Hour of Day",
       y = "Active Event Count")
```
