---
title: "Create Calendar Table"
author: "Donnie Minnick"
date: "`r Sys.Date()`"
output:
  md_document:
    variant: markdown_github
---

## Problem

You need to create a calendar dimension table to support a data analysis or business intelligence project.

## Solution

#### Load Libraries

This solution uses the `dplyr`, `knitr`, `lubridate` and `stringr` libraries.  The code chunk below checks for installation and, if needed, will install the libraries and load them.  The `ggplot` library is used to generate the plot at the end of this example.

```{r load_libraries, message = FALSE}
packages <- c("dplyr", "knitr", "lubridate", "stringr")

installed_packages <- packages %in% rownames(installed.packages())

if(any(installed_packages == FALSE)) {
  
  install.packages(packages[!installed_packages])
  
}

invisible(lapply(packages, library, character.only = TRUE))
```

#### Set Date Range

Set the start and end dates for the table.

```{r set_date_range, message = FALSE}
start_date <- as.Date("2022-01-01")
end_date <- as.Date("2022-12-31")
```


#### Create Data Frame

Create the data frame using the date sequence.

```{r create_data_frame, message = FALSE}
calendar <- data.frame(date = seq.Date(start_date, 
                                       end_date, 
                                       by = "1 day"))
```

#### Add Variables

Add date-related variables to the data frame.

```{r add_variables, message = FALSE}
calendar <- calendar %>%
  mutate(year = year(date),
         quarter = quarter(date),
         month = month(date),
         week = week(date),
         day_of_month = day(date),
         year_and_quarter = paste(year(date), 
                                  "-",
                                  str_pad(quarter(date), 2, "left", "0"),
                                  sep = ""),
         year_and_month = paste(year(date), 
                                "-",
                                str_pad(month(date), 2, "left", "0"),
                                sep = ""),
         month_long_name = month(date, TRUE, FALSE),
         month_short_name = month(date, TRUE, TRUE),
         day = wday(date),
         day_long_name = wday(date, TRUE, FALSE),
         day_short_name = wday(date, TRUE, TRUE))
```

Here's the first ten rows of the table.

```{r view_table, message = FALSE}
kable(head(calendar, 10),
      col.names = c("Date", "Year", "Quarter", "Week",
                    "Month", "Day of Month", "Year and Quarter", 
                    "Year and Month", "Month Long", "Month Short", 
                    "Day", "Day Long", "Day Short"),
      caption = "Calendar Table",
      align = c("l", "r", "r", "r",
                "r", "r", "r", 
                "r", "r", "r",
                "r", "r", "r"))
```

## Discussion


