---
title: "Likert Diverging Stacked Bar Chart"
author: "Donnie Minnick"
date: "`r Sys.Date()`"
output:
  md_document:
    variant: markdown_github
---

## Problem

When you need to view Likert data using a diverging stacked bar chart.

## Solution

A Likert scale is a psychometric scale named after its inventor, American social psychologist Rensis Likert, which is commonly used in research questionnaires.  It is the most widely used approach to scaling responses in survey research.

This solution is based on a five-point Likert scale.

#### Load Libraries

This solution uses the `dplyr`, `ggplot2`, `ggstats` and `knitr` libraries.  The code chunk below checks for installation and, if needed, will install the libraries and load them.

```{r load_libraries, message = FALSE}
packages <- c("dplyr", "ggplot2", "ggstats", "knitr")

installed_packages <- packages %in% rownames(installed.packages())

if(any(installed_packages == FALSE)) {
  
  install.packages(packages[!installed_packages])
  
}

invisible(lapply(packages, library, character.only = TRUE))
```

#### Create Likert Data

Create a data frame, generating one hundred random responses for three questions using a five-point Likert scale.

```{r create_likert_data, messages = FALSE}
set.seed(91210)

likert_scale <- c("Strongly Agree", 
                  "Agree", 
                  "Neutral", 
                  "Disagree", 
                  "Strongly Disagree")

q1 <- sample(likert_scale, 100, replace = TRUE)

q2 <- sample(likert_scale, 100, replace = TRUE)

q3 <- sample(likert_scale, 100, replace = TRUE)

responses <- data.frame(cbind(q1, q2, q3))
```

Here's what the data frame looks like.

```{r view_data_frame, messages = FALSE}
head(responses, 10)
```

#### Generate Chart

Generate a diverging stacked bar chart with the data frame.  This chart type is effective for quickly understanding the distribution of responses on the five-point scale.  Note that the neutral category is centered in the chart and both negative and positive responses extend from the center.

```{r generate_chart, messages = FALSE}
gglikert(responses) +
    labs(title = "Likert Diverging Stacked Bar Chart",
         subtitle = "This is a Likert example with minimal formatting.",
         x = "Percent",
         y = "Question")
```

#### Customize Chart

Likert charts generated with `gglikert` can be customized via a range of arguments as well as functions from the `ggplot2` package that allow for customization of colors, for example.

```{r generate_customized_chart, messages = FALSE}
gglikert(responses,
         labels_size = 3,
         labels_accuracy = 1,
         labels_hide_below = .2,
         labels_color = "black") +
    scale_fill_brewer(palette = "RdYlBu") +
    labs(title = "Customized Likert Chart",
         subtitle = "This is a Likert example with some formatting.",
         x = "Percent",
         y = "Question")
```

## Discussion

Diverging stacked bar charts are an effective means to visualize survey data when collected via a Likert scale.

I've also used this chart type when showing relative maturity levels on a five-point scale for capability assessments.
