---
title: "Sankey Chart"
author: "Donnie Minnick"
date: "`r Sys.Date()`"
output:
  md_document:
    variant: markdown_github
---

## Problem

You need to visualize the flow of resources or information within a system.

## Solution

A sankey chart is a type of flow diagram that visualizes the flow of information within a system through a series of interconnected pathways. 

The width of the arrows or lines in a sankey chart is proportional to the quantity of the flow, allowing for a clear representation of the relative magnitudes of different components.

These charts are particularly useful for presenting complex systems with multiple inputs, outputs, and interconnections in a visually intuitive and easily understandable manner.

#### Load Libraries

This solution uses the `dplyr`,`knitr` and `networkD3` libraries.  The code chunk below checks for installation and, if needed, will install the libraries and load them.

```{r load_libraries, message = FALSE}
packages <- c("dplyr", "knitr", "networkD3")

installed_packages <- packages %in% rownames(installed.packages())

if(any(installed_packages == FALSE)) {
  
  install.packages(packages[!installed_packages])
  
}

invisible(lapply(packages, library, character.only = TRUE))
```

#### Create Sample Data

```{r create_nodes, message= FALSE}
nodes <- read.csv("../inst/extdata/sankey/nodes.txt",
                   header = TRUE,
                   sep = ",",
                   na.strings = c(""),
                   stringsAsFactors = FALSE)

kable(nodes,
      col.names = c("Id", "Name"),
      caption = "Nodes",
      format.args = list(big.mark = ","),
      align = c("l", "l"))
```



```{r create_sample, message = FALSE}
sample <- read.csv("../inst/extdata/sankey/sample.txt",
                   header = TRUE,
                   sep = ",",
                   na.strings = c(""),
                   stringsAsFactors = FALSE)

kable(head(sample, 10),
      col.names = c("Position", "Starting Status", "Ending Status"),
      caption = "Job Search Process Flow",
      format.args = list(big.mark = ","),
      align = c("l", "l", "l"))
```



```{r}
sample <- sample %>%
  group_by(starting_status,
           ending_status) %>%
  summarise(count = n())

kable(sample,
      col.names = c("Starting Status", "Ending Status", "Count"),
      caption = "Job Search Process Flow",
      format.args = list(big.mark = ","),
      align = c("l", "l", "r"))
```



```{r}
sample$source_id <- match(sample$starting_status, nodes$node_title) - 1
```



```{r}
sample$target_id <- match(sample$ending_status, nodes$node_title) - 1
```

Generate the sankey chart using the summarized sample data.

```{r}
p <- sankeyNetwork(Links = sample,
                   Nodes = nodes,
                   Source = "source_id",
                   Target = "target_id",
                   Value = "count",
                   NodeID = "node_title",
                   sinksRight = FALSE,
                   iterations = 0,
                   fontFamily = "Calibri",
                   fontSize = 11); p
```



## Discussion


